Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-11 image_analysis_paper_development2/paper_figures/fig3.tex
added compiled figures

\documentclass[twocolumn, 10pt]{article}

\usepackage[margin=0.75in]{geometry}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{graphicx}

\usepackage{color}

\usepackage{epstopdf}

\usepackage{natbib}

\renewcommand{\figurename}{Fig.}
%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
\newcommand{\SI}[0]{supplementary material}
\newcommand{\fig}[0]{Fig.}

\hyphenpenalty=10000

\makeatletter
\newcommand{\customlabel}[2]{%
\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
\makeatother

\graphicspath{{paper_figures/}}

\customlabel{fig:DV_view}{S1}
\customlabel{fig:1d_example}{S2}
\customlabel{fig:eigenvalues_drosophila_live}{S3}
\customlabel{fig:eigenvalues_zebrafish}{S4}
\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
\customlabel{fig:eigenvalues_wing_disc}{S6}
\customlabel{fig:PCA_compare}{S7}

\title{Temporal ordering and registration of images in studies of developmental dynamics}

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1], \\
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,and
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}

\date{}

@@

\maketitle

\begin{abstract}

Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide software to aid in the application of our methodology to other experimental data sets.
\vspace{1em}

\noindent
KEY WORDS: temporal ordering, image registration, vector diffusion maps

\end{abstract}

\section*{Introduction}

\begin{figure}[t]
\includegraphics{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. (A) Images of ``samples'', each in a different orientation and a different stage of development. (B) Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
\end{figure}

In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
%
The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
%
Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known.
%
Typically, what is known is
a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
%
Furthermore, images are often collected in different spatial orientations.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
%
%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

Temporal ordering and registration of images can be done manually
when the number of images is small and the differences between them are visually apparent.
%
\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
%
In contrast to this example, real data pose nontrivial challenges for both registration and temporal ordering.
%
In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
%
Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.

We present a robust algorithmic approach to simultaneous registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of landmarks for registration or markers of developmental progression.
%
The approach is based on vector diffusion maps \citep{singer2012vector}, a manifold learning algorithm which simultaneously addresses the problems of registration and temporal ordering.
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.

Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four experimental data sets.
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
%
Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
%
We also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

\section*{Results}

\subsection*{Vector diffusion maps for registration and temporal ordering}

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries, such as rotations of the images, which one would like to factor out,
and ``additional" directions of variability, such as temporal dynamics, which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parametrizing the curve.}
\label{fig:schematics}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
%
However, pairwise alignments can also be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to test our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

\subsection*{Method validation using live imaging}

\subsubsection*{{\em Drosophila} gastrulation}

\begin{figure*}[t]
\raisebox{0.75cm}{\textsf{A}}
\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\
2 & 4.98$^{\circ}$ & 0.9994 \\
3 & 6.47$^{\circ}$ & 0.9795 \\
4 & 7.03$^{\circ}$ & 0.9953 \\
5 & 3.04$^{\circ}$ & 0.9959 \\
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Scale bar indicates 50$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A registered and ordered by vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

To validate the proposed approach, we first applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
%
Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
%
Each frame was arbitrarily rotated, and the order of the frames was scrambled.
%
The task is now to register these images and order them in time to reconstruct the developmental trajectory.

We used vector diffusion maps to register and order the images.
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics.
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and rank orders, respectively, for the entire data set.
%
Both the angles and the ranks are recovered with a high degree of accuracy.
%
We note that determining which end of the trajectory corresponds to early in the developmental progression is a post-processing task that requires some {\em a priori} information.

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
%In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
%At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

\subsubsection*{Zebrafish epiboly}

\begin{figure*}[t]
\begin{minipage}[c]{13.75cm}
\vspace{0cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=13cm]{zebrafish_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=13cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\begin{minipage}[c]{3.75cm}

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Scale bar indicates 200$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \citep{kane1996zebrafish}.
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the two-dimensional frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
%
We then used vector diffusion maps to register and order the frames.
%
The results are shown in \fig~\ref{fig:zebrafish}B.
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
%
Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
%
%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

\subsection*{Data sets with intersample variability}

\subsubsection*{Fixed images of {\em Drosophila} gastrulation}

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
%
In practice, we are interested in cases where each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such variations,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
%
Registered and ordered images of individual embryos can then be used to construct a representative average trajectory. 
%
%However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
%To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set (see \SI).
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

\begin{figure*}[t!]
\begin{minipage}[b]{12cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).  Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

From this average trajectory, we can now easily see the developmental progression consistent with the known dynamics:
%
dpERK first appears as two lateral peaks at the ventrolateral side of the embryo, and a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks and without {\it a priori} knowledge of developmental features.
%

To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system.
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo.
%
We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images.
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering.
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{fig:drosophila_fixed_images}B , and the rank correlation (see \fig~\ref{fig:drosophila_fixed_images}D) shows that our ordering is consistent with the expert ordering.

\subsubsection*{Fixed z-stacks of {\em Drosophila} wing discs}

\begin{figure*}[t]
\raisebox{5.5cm}{\textsf{A}}
\includegraphics[width=6cm]{wing_disc_example}
%
\hfill
%
\raisebox{5.5cm}{\textsf{B}}
\includegraphics[width=10cm]{wing_disc_ordered}

\vspace{0.25cm}
\raisebox{2.25cm}{\textsf{C}}
\includegraphics[width=17.5cm]{wing_disc_average}

\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
%
%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
%
%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.

In this section we show that the approach can readily be applied to three-dimensional data. 
%
We restrict ourselves to the case where an obvious fixed axis exists, so that only rotations of the three-dimensional data around this axis need be taken into account.  
%
This does not constitute an inherent limitation for vector diffusion maps. 
%
While for simplicity here we will not discuss the general  case, incorporating general 3D symmetries is possible \citep{arie2012global, wang2013exact, cucuringu2012eigenvector}.

%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
%
%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.

To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs \citep{hamaratoglu2011dpp}.
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs during metamorphosis.
%
The wing disc is an imaginal disc that turns into a wing.
%
The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization.
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals.
%
The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the factors that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
%
%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
%
Therefore, we need not consider registration in all three dimensions, and can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
%
To register the data, we first aligned the maximum intensity projections using angular synchronization.
%
We then used these rotations to register the full three-dimensional data in the x-y plane.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
%
Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
%
%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now clearly see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

%The vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
%
%In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions.
%
%The vector diffusion maps algorithm then, instead of revealing the temporal order, reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object.
%
%This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.

\subsection*{Computational requirements}

\begin{figure*}[t]
\raisebox{5cm}{\textsf{A}}
\includegraphics[width=5.5cm]{timing_studies_nimages}
%
\raisebox{5cm}{\textsf{B}}
\includegraphics[width=5.5cm]{timing_studies_npixels}
%
\raisebox{5cm}{\textsf{C}}
\includegraphics[width=5.5cm]{timing_studies_nrot}

>>>>

\documentclass[10pt]{article}

\usepackage[paperwidth=18cm, paperheight=8cm, margin=0cm]{geometry}
\usepackage{graphicx}
\hyphenpenalty=10000

@@

\raisebox{0.75cm}{\textsf{A}}
\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\
2 & 4.98$^{\circ}$ & 0.9994 \\
3 & 6.47$^{\circ}$ & 0.9795 \\
4 & 7.03$^{\circ}$ & 0.9953 \\
5 & 3.04$^{\circ}$ & 0.9959 \\
 \hline
\end{tabular}}}}
